---
title: "GDP Growth and Anti-Government Protesting"
author: "Ines de la Morena"
description: "What is the relationship between the GDP growth rate and anti-government demonstrations worldwide?"
output:
  distill::distill_article:
    self_contained: false
---

Project Introduction
Introduction: describe the research question and main hypothesis; describe why it is important. (1-2 paragraphs) (2pts)


What is the relationship between the GDP growth rate and anti-government demonstrations? I believe that if GDP growth is stagnant or negative, there would be a rise in anti-government demonstrations. In "How Democracies Die" by Ziblatt and Levitsky, they stated that the better the economic state of a nation, the less likely there is to be democratic breakdown or civil unrest. If this is true, then I am curious to see if GDP growth over time increases or decreases national unity and discontent, using anti-government demonstrations as a sort of proxy. The data set I will be using is the political violence data set provided on the course website. Thus, my explanatory variable is the real GDP growth rate. The outcome that I am interested in exploring based on variations in GDP growth is anti-government demonstrations. It is measured in the number of anti-government demonstrations in 1960, 1970 and 1980, by country. To support my hypothesis, two relationships could be true. First, a negative or stagnant growth of GDP coupled with a rise in anti-government demonstrations. Alternatively, a positive GDP growth and low number of anti-government demonstrations. A pattern of growth coupled with a rise in anti-government demonstrations or sentiment would disprove my hypothesis, as well as its inverse. My regression coefficient, when controling for initial GDP, should thus be negative. I am interested in exploring this relationship, or something similar to it, in order to understand politics and my own Spanish background better. 


Data 
Data used: violence.csv

This data is available on the course website. It consists of 208 rows, one for each of the countries in the dataframe, and over a hundred variables. The variables used for this project are the gyp (representing Growth rate of real per capita GDP), lrgdp (Log of initial Income: Log of real per capita GDP), and agovdem (Anti-government demonstrations: Any peaceful public gathering of at least 100 persons). The GDP and GDP growth data are from the Bureau of Economic Analysis. They each have the numbers 60, 70, or 80 as a suffix (representing the years 1960, 1970, and 1980, respectively). While I originally wished to do coups, coups were few and far between in the data set, and also set a high threshold for exemplifying government unrest. 

My research design is technically a longitudinal study that aims to pinpoint the relationship between GDP growth and anti-government protests. It takes into account three distinct points in time: 1960, 1970, and 1980. However, in order to reduce noise, the three variables were averaged across those three points in time. Although this gives me one value for each variable, it is not a study of a single distinct point in time and thus it can't accurately be described as a cross-sectional study. The data set-up process and the regression are shown below:   

```{r}
library(dplyr)
library(tidyverse)
violence <- read_csv('violence.csv')

violence <- subset(violence, select = c(sftgname, agovdem60, gyp60, lrgdp60, agovdem70, gyp70, lrgdp70, agovdem80, gyp80, lrgdp80))

violence <- violence |>
 mutate(mean_protest = (agovdem60 + agovdem70 + agovdem80)/3,
        mean_growth = (gyp60 + gyp70 + gyp80)/3,
        mean_gdp = (lrgdp60 + lrgdp70 + lrgdp80)/3)

regression_multiyear = lm(mean_protest ~  mean_growth + mean_gdp,
               data = violence)
regression_multiyear
```

When looking at the data, it was clear to me that it was non-linear, and its shape resembled more a quadratic function. I thus used the loess method to visualize the regression I ran in the previous section in order to best fit the data. For the visualization of my outcome of interest, I expected a downward trending line. This would signify that as the GDP growth rate increased, there would be less anti-government demonstrations, signifying content for the ruling party. Although my data had few values with negative GDP growth, I expected said values to have high positive correlations with anti-government demonstrations. If the country was not only not improving a lot but worsening, I would think that more people would demand some change in the government through anti-government demonstrations. In order to see if my hypothesis was true, I visualized the data as follows: 


```{r}

scatter <- regression_multiyear |>
  ggplot(aes(x = mean_growth, y = mean_protest)) +
  geom_point(color = "deeppink", alpha = 0.25) +
  geom_smooth(method= 'loess') +
  labs(
    x = "Average GDP Growth Rate",
    y = "Average Anti-government demonstrations, more than 100 people",
    title = "GDP Growth Rate and Demostrations Worldwide in the 60s, 70s, and 80s"
  ) 
scatter

```

Results 
Results section: plot of main analysis + regression output + 2-3 paragraphs of description and interpretation of the plots and regression (including interpreting the main coefficient of interest and describing if it is statistically significant and if we should interpret it causally). This section could be longer if you choose to include additional analyses. (8pts). For the results section, you should interpret (in plain English) the main coefficient of interest in your regression. You should also comment on the statistical significance of the estimated coefficient and whether or not you believe the coefficient to represent a causal effect.

My data visualization surprised me. The results of my regression show an upward-sloping trend, with a peak at around 0.020 GDP growth rate. I looked back into the data and identified a few outliers, such as Uruguay and the United States in 1960. That year, there were around 20 anti-government protests in each country, in Uruguay due to growing stagflation and looming threat of dictatorship, and in the United States due to the Vietnam War. Regardless of these outliers and some others, the correlation peaks at around a 0.015 growth rate, or 1.5%. At this level, there is a likelihood of about 2 anti-government protests in a given year. As the GDP growth rate grows to 2% per year or more, there is a lower correlation with anti-government protests. At growth rates of 4% or more, there is not enough data to definitively state, as the graph would suggest, that anti-government protests would increase. The upward slope is mostly based on outliers such as Cyprus. In order to better understand this relationship, the regression coefficients and p-values were calculated as follows: 

```{r}

summary(regression_multiyear)

regression <- regression_multiyear |> 
  broom::tidy() |>
  select(term, estimate, p.value) |>  
  knitr::kable(digits = 2)
regression

```

Could the variable I controled for be a better predictor??

It seems like most data points are concentrated alomg the 0-0.05 GDP growth rate. For those on the higher extremes of GDP growth, there are few data points but they all seem to be closer to zero in the sep_gdp histogram, with a few exceptions/outliers. The second graph also as I said has a lot values at zero, and all are between 0-3. Those with repeated illegal or forced changes in government seem to be more concentrated in the low end of GDP growth, which could confirm my hypothesis that lowe levels of economic growth lead to more dissatisfaction with the ruling party. But, I would have to do more analyses to determine the significance. I think also for the second graph it would be good to perhaps represent it in another way such as a histogram or box plot. 

Conclusion 
Conclusion section: 1 paragraph (i) summarizing results and assessing the extent to which you find support for your hypothesis; (ii) describing limitations of the analysis and threats to inference (missing data, confounding, etc), and stating how you could improve your analysis if you had more time/money. (2pts)

















